Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-30 paper.tex
(>) figures placement
...

\abstract{TODO.

\smallskip

\noindent \textbf{Keywords:} TODO, TODO, TODO.
}

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control-flow mechanisms, such
      as \code{await} (to suspend a line of execution), and \code{par} (to
      combine multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution.
      There is no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU provides primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

@@

\begin{figure*}[th!]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Bomber::Bomber (Pingu* p) :
  <...>
  sprite(<...>),          // bomber sprite
  sound_played(false),    // tracks state 2
  particle_thrown(false), // tracks state 3
  colmap_exploded(false), // tracks state 3
  gfx_exploded(false)     // tracks state 4
{
  <...>
  // 1. plays a "Oh no!" sound.
  play_sound("ohno");
}

@@

The equivalent code in \CEU for the \emph{Bomber action} in
Figure~\ref{lst.bomber}.b does not rely on state variables and reflects the
sequential state machine implicitly, using \code{await} statements to separate
the effects in direct style.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code/await} abstractions and execute in separate through the \code{spawn}
primitive (ln. 4--5).
The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every game frame.
%
The code tracks the animation aliveness (ln. 7--27) and, on termination,
performs the last bomber effect, killing the pingu (ln. 30).
As soon as the animation starts, the code performs the first effect (ln. 9).
The intermediate effects are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all effects occur in a contiguous
chunk of code (ln. 7--30), which handles no extra functionality.

@@

@@

\begin{figure*}[t]
\centering
\includegraphics[width=\textwidth]{continuation}
\caption{ Continuation (C++) vs Direct (\CEU) Styles.
\label{fig.story}
}
\end{figure*}

@@

TODO: non reactive, C++ integration
- TODO: OO state + methods
- eliminar estados explicitos com estruturas de controle apropriadas

We promote the \emph{structured synchronous reactive} programming model of
\CEU for the development of games.
We present in-depth use cases categorized in four control-flow patterns applied
to \emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to
other games.

We show how the standard way to program games with objects and callbacks in C++
hinders structured programming techniques, such as support for sequential
execution, long-lasting loops, and persisting local variables.
In this sense, callbacks actually disrupt structured programming, becoming
["our generation’s goto"][goto] according to Miguel de Icaza.

Overall, we believe that most difficulties in implementing control behavior in 
game logic is not inherent to this domain, but a result of accidental
complexity due to the lack of structured abstractions and an appropriate
concurrency model to handle event-based applications.

TODO: rever summaries por advantage qualitativa vs LoCs

[goto]: tirania.org/blog/archive/2013/Aug-15.html

@@

We would like to thank Leonardo Kaplan and Alexander Tkachov for early
explorations and prototypes of the game rewrite.

>>>>

\abstract{
We present a case study of rewriting the video game Pingus from C++ to the
structured synchronous reactive language \CEU.
%
\CEU supports reactive control-flow primitives that eliminate callbacks and let
programmers write code in direct and sequential style.
Structured reactivity helps describing complex control-flow relationships in
the game logic more concisely.
%
We demonstrate gains in productivity for six behaviors in the game logic of
Pingus through a qualitative analysis of the proposed solutions in \CEU in
comparison to the original implementations in C++.
%
We also categorize the behaviors in four recurrent control-flow patterns that
likely apply to other games.
\smallskip

%\noindent \textbf{Keywords:} \CEU, Game Logic, Event-Driven Programming.
\noindent \textbf{Keywords:} C++, \CEU, Control Flow, Event-Driven Programming, Game Logic, Synchronous Reactive Programming.
}

@@

Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
to handle events from the environment (e.g., key presses and timers) and also
as a notification mechanism between entities in the game logic.
%
The observers are short-lived callbacks that have to execute as fast as
possible to keep the game reactive to incoming events in real time.
%
For this reason, callbacks cannot use long-lasting locals and loops, which are
elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \emph{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{tirania.org/blog/archive/2013/Aug-15.html}}
%\footnote{``Escape from Callback Hell'':
%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

In this work, we advocate structured synchronous reactive programming as a more
productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control-flow mechanisms, such
      as \code{spawn} and \code{await} (to create and suspend an activity).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      activity.
      There is no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
%\CEU provides primitives that help describing complex control-flow
%relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

@@

\begin{figure*}[!t]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
Bomber::Bomber (Pingu* p) :
  <...>
  sprite(<...>),          // bomber sprite
  sound_played(false),    // tracks state 2
  particle_thrown(false), // tracks state 3
  colmap_exploded(false), // tracks state 3
  gfx_exploded(false)     // tracks state 4
{
  <...>
  // 1. plays a "Oh no!" sound.
  play_sound("ohno");
}

@@

The equivalent code in \CEU for the \emph{Bomber action} in
Figure~\ref{lst.bomber}.b does not rely on state variables and reflects the
sequential state machine implicitly, using \code{await} statements to separate
the effects in direct style.
%
The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
retains runtime state, such as local variables and the program counter, across
reactions to events (i.e., across \code{await} statements).
The pingu movement and sprite animation are isolated in two other
\code{code/await} abstractions and execute in separate through the \code{spawn}
primitive (ln. 4--5).
In \CEU, if multiple abstractions react to the same event, the scheduler
employs lexical order to preserve determinism, i.e., the \code{spawn} that
appears first in the source code executes first~\cite{ceu.mod15}.
The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
callback of C++ and occurs on every game frame.
%
The code tracks the animation aliveness (ln. 7--27) and, on termination,
performs the last bomber effect, killing the pingu (ln. 30).
As soon as the animation starts, the code performs the first effect (ln. 9).
The intermediate effects are performed when the corresponding conditions occur
(ln. 12,16,24).
The \code{do-end} block (ln. 19--25), restricts the lifespan of the
single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
block terminates and automatically destroys the spawned abstraction (removing
it from the screen).
%
In contrast with the implementation in C++, all effects occur in a contiguous
chunk of code (ln. 7--30), which handles no extra functionality.

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=\textwidth]{continuation}
\caption{ Continuation (C++) vs Direct (\CEU) Styles.
\label{fig.story}
}
\end{figure*}

@@

@@

We advocate \emph{Structured Synchronous Reactive Programming} as a productive
alternative for game logic development.
%
We compare the implementation of six game behaviors in C++ and \CEU and discuss
how structured reactive mechanisms can eliminate callbacks and let programmers
write code in direct style.

We categorize the behaviors in four recurrent control-flow patterns:
%
\emph{State machines} are the workhorses of the game logic, appearing in
animations, AI behaviors, and input handling.
\CEU can encode states implicitly with sequential statements in the same
contiguous block, eliminating shared state variables and improving code
encapsulation.
%
\emph{Continuation passing} controls the overall structure of the game, such as
screen transitions and level progressions.
Similarly to state machines, \CEU can describe the flow of the game with
sequential statements in a self-contained block, eliminating explicit
data structures and continuation variables (e.g., stacks and boolean flags)
spread across multiple classes.
%decouple activities by
%
\emph{Dispatching hierarchies} spread input events through the game entities
and serve as a broadcast communication mechanism.
Event broadcasting is at the core of the semantics of \CEU, allowing entities
to react directly to inputs and bypass the program hierarchy entirely.
In C++, a considerable amount of interleavning code in the game logic and
engine is just destinated to event dispatching.
%
\emph{Lifespan hierarchies} manage the memory and visibility of game entities
through class fields and containers.
In \CEU, all entities have an associated lexical scope, similarly to local
variables with automatic memory management.
This reduces explicit manipulations of pointers and references considerably.

Overall, we believe that most difficulties in implementing control-flow
behavior in game logic is not inherent to this domain, but a result of
accidental complexity due to the lack of structured abstractions and an
appropriate concurrency model to handle event-based applications.

@@

The author would like to thank Leonardo Kaplan and Alexander Tkachov for early
explorations and prototypes of the game rewrite.
